### 2014-09-06 eval.tex
-xz

\section{Evaluation}~\label{sec:eval}
Our evaluation focuses on the effectiveness  and scalability of our approach.
To measure the effectiveness, we compare with the predictive analysis, {\sf RV}~\cite{}. We chose RV for two main reasons: (1) {\sf RV} is the only open source predictive analysis, (2) {\sf RV} represents the state-of-the-art approach, which is theoretically proven to have higher detection capability than other approaches.

{\bf Prototype Implementation}  In our prototype implementation of \tool, we used the Z3 SMT solver for constraint resolution \cite{MouraB08}. For static analysis, to discover unexecuted branches, we utilize the SOOT compiler framework \cite{Vallee-RaiCGHLS99}.

@@

To address the scalability limitations of the constraint solver (Z3 in our case), we again apply a splitting optimization. The full trace is split into $N$ equal-length subtraces or ``time windows'' (in our experiments, $N=1,000$). The values of local and shared variables are then passed at the end of a given window as inputs to the next window. For lack of space, we omit the concrete details of this optimization, and instead refer the reader to Huang et al. \cite{HuangMR14}, who apply a similar optimization.

{\bf Evaluation Method} We conducted our experiments on a set of 16 applications, which were also used to evaluate {\sf RV}. Specifically, the set includes large applications such as {\sf Jigsaw}, {\sf Xalan} and {\sf Lusearch}. To handle large applications that make use of reflection, we applied the {\sf Tamiflex} tool~\cite{}, which replaces reflective calls with the concrete method invocations recorded in the observed run for the purpose of effective static analysis.  We omit the benchmark {\sf eclipse} because the current version of {\sf Tamiflex}  leads to abnormal execution after the instrumentation, which throws an exception when the main service is started. By applying our tool to such abnormal execution, we identify only 3 races, similarly to the report of RV~\cite{}.
In addition, the {\sf montecarlo} benchmark requires an external input that we downloaded from the internet and simplified. For the large applications, we utilized the most lightweight available configuration.

 As the detection capability  depends on the observed run, for fair comparison we monitor the execution once by recording all necessary information required by both approaches, and then apply both techniques to the monitored run. Besides, our reported data for RV may be different from the original report for two reasons: (1) the different observed runs lead to different sets of races, (2) the original implementation of RV contains a bug in identification of the branches, which leads to  misses of many branches and  incorrect reduction of dependences during the analysis. We confirmed the bug with the author and fixed the bug in our experiments.\footnote{We have created an in-depth report on the bug, along with a witness test case. (See:  \url{https://sites.google.com/site/recipe3141/}.)}

Our experiments and measurements 
were all conducted on an x86 64 Thinkpad W530 worksta- 
tion with eight 2.30GHz Intel Core i7-3610QM processors, 
16GB of RAM and 6M caches. The workstation runs version 12.04 of the Ubuntu Linux distribution, and has the Sun 
64-Bit 1.6.0\_26 Java virtual machine (JVM) installed.
%\footnote{We note that Sun JDK 1.7 does not support the transformation of dacapo applications with {\sf tamiflex}.}

@@

The Trace section includes the number of threads ($Th$), the number of shared reads ($Reads$) and shared writes ($Writes$), the subset of shared reads that read the base object references  ($Base$), where the  base object references are references used as the base/target in the following field reference or the method invocation,  the number of branches ($Br$), the synchronization events ($Sync$) and the total number of events ($Total$), which includes local accesses in addition to the aforementioned events. Specifically, branch events are reported in the form $A/B$, where $A$ refers to the number of branches used by our analysis and $B$ refers to the number of branches used by RV. To ensure that the predicted run sees the same base object at each shared read/write, {\sf RV} inserts the artificial branch immediately in front of each shared access (and array accesses). We do not use such artificial branches.

We make interesting observations about the Trace section. The non-local events, i.e., all the events listed in Table~\ref{tab:main}, occupy around 30\% of the total trace in the first 7 benchmarks, but occupy less than 1\% of the trace in the rest of the benchmarks, which have relatively more complex logic. Reads of the shared base objects occupy a small portion ($\frac{1}{10}$-$\frac{1}{3}$) of the total shared reads. The rest of the shared reads read only primitive values or 
references not involved in field accesses (e.g., references involved in nullness checking). Besides, our analysis involves significantly less branch events as compared to the {\sf RV} approach. This difference plays an important role in the detection, which we will explain shortly.

@@

First of all, {\sf Recipe} finds $>$150 more races than {\sf RV}. There are several reasons for that: (1) {\sf Recipe} can reason about the accesses in the un-executed paths, while {\sf RV} and {\sf Recipe-s} can only reason about the accesses in the executed paths. (2) {\sf Recipe} or {\sf Recipe-s} relax the scheduling even if it breaks the inter-thread read/write dependence structure in the observed run. {\sf Recipe} allows the majority of the shared reads, i.e., the reads of primitive values, to freely read a different value from a different write as long as the value leads to the same branch decisions. {\sf RV}, however, requires them to read the same values as in the observed run. (3) An critical optimization proposed by {\sf RV} is to preserve dependences only for the reads before the preceding branches  of the racy events, rather than all the reads. However, this optimization is underplayed by the fact that {\sf RV} introduces huge amount of artificial branches, i.e., one before each shared field access, which ensures the use of the same base objects.
We get rid of such artificial branches and instead, rely on the small amount of base read events to ensure the use of the same base objects (Section~\ref{}). Our strategy reduces the number of branches greatly and amplifies the effectiveness of the optimization. We also carried out case studies (Section~\ref{}) to better illustrate the scenarios.

Another interesting observation is that although {\sf Recipe} should produce all races found by {\sf RV} in theory, {\sf Recipe} may miss races found by {\sf RV} in practice (Column $Diff'$), i.e., {\sf Recipe} is not strictly more effective than {\sf RV} in practice. The reason for this is limitations of the underlying constraint solver: (1) the solver cannot compute constraints with very complex arithmatic operations, and (2) the solver does not support some program constants such as the scientific notation $3E-10$. In our empirical evaluation, we encounted only one benchmark, account, where these issues manifested.

The last section, Running time, compares the analysis time for both approaches. We find our approach is significantly slower than {\sf RV}, e.g., {\sf RV} often finishes within 200 seconds, while our approach may take more than 1 hour. This is because our approach needs to reason about the computation among variables inside the local access events, while {\sf RV} needs to only reason about the order relations among the events.

@@

{\bf Relaxing Inter-thread dependencies\ } Figure~\ref{fig:relax1} demonstrates the relaxation of inter-thread dependence enabled by our approach. 
The code is from the benchmark {\sf bbuffer}, where the line number is marked. Huang et al.~\cite{} detect the race between lines 291 and 400, but fail to detect the race between lines 294 and 400. The reason is that in the observed run, the execution follows the order 400, 291 and 294. For the event at line 294, its preceding branch at line 291 reads from line 400. Therefore, Huang et al.~\cite{}
 require the predicted run to preserve the dependence between line 291 and line 400, so that line 291 reads exactly the same value and the branching expression evaluates to the same truth value. The dependence enforces the order constraint, $400 \rightarrow 291$, which further enforces the order $400\rightarrow 291 \rightarrow 294$.    Our approach does not require the existence of such dependence. Specifically, we allow line 291 to happen before line 400 in the predicted run as long as the value read by it leads to the same branch decision, which is true in this case.  As a result, there is no order constraint between line 294 and line 400, and the two form a race. We were able to replay this race  easily using the eclipse IDE breakpoints.

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{cases/Visio-bbuffer.pdf}
\caption{Relaxation of Inter-thread dependence}\label{fig:relax1}
\end{figure}

{\bf Relaxing the Paths\ } Figure~\ref{fig:relax2} demonstrates how our approach relaxes the scheduling to account for  unexecuted paths. 
All threads invoke the method {\sf Sorting}, which recursively starts two child threads if there are two more available entries in the thread pool (lines 8-11), or  starts one child thread if there is only one available entry (lines 4-5).  The availability is computed through the static method {\sf available} with the constant {\tt total}, which is equal to 5. The shared variable {\tt alive} denotes the used entries.

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{cases/Visio-msort.pdf}
\caption{Relaxation of Paths}\label{fig:relax2}
\end{figure}

\section{Threats to Validity}
A Java method can consist of up to 65535 bytecode instructions. Therefore, we count the number of bytecode instructions inside each method. If the resulting number exceeds 65525, we avoid instrumenting the method. The consequence is that we will potentially races inside a method for which we skipped instrumentation.. In this case, we specify the variables read from the method to be equal to their concrete values in the constraints. The constraint solver can proceed safely without being affected by such methods.

A second point regards support for certain operators. In our current implementation, we do not support the boolean operators, including e.g. $\&$, as well as the bit operator $<<$, which accounts for most of our misses.

>>>>

\section{Evaluation}~\label{sec:eval}
Our evaluation focuses on the effectiveness  and scalability.
To measure the effectiveness, we compare with the predictive analysis, 
{\sf RV}~\cite{}. We chose RV for two main reasons: (1) {\sf RV} is the 
only open source predictive analysis, (2) {\sf RV} represents the 
state-of-the-art, which is theoretically proven to have higher detection 
capability than other approaches.

{\bf Prototype Implementation}  We use the Z3 SMT solver for constraint 
resolution \cite{MouraB08}. For static analysis, to discover unexecuted 
branches, we utilize the SOOT compiler framework \cite{Vallee-RaiCGHLS99}.

@@

To improve the scalability in constraint solving, we again apply 
a splitting optimization. The full trace is split into $N$ equal-length 
subtraces or ``time windows'' (in our experiments, $N=1,000$). The values 
of local and shared variables are then passed at the end of a 
given window as the inputs to the next window. For lack of space, we 
omit the details, and instead refer the reader to 
Huang et al. \cite{HuangMR14}, who apply a similar optimization.

{\bf Evaluation Method} We conducted our experiments on a set of 16 
applications, which were also used to evaluate {\sf RV}. Specifically, 
the set includes large applications such as {\sf Jigsaw}, 
{\sf Xalan} and {\sf Lusearch}. To handle large applications that make 
use of reflection, we applied the {\sf Tamiflex} tool~\cite{}, which 
replaces reflective calls with the concrete method invocations recorded 
in the observed run.
% for the purpose of effective static analysis.  
We omit the benchmark {\sf eclipse} because the current version 
of {\sf Tamiflex}  leads to abnormal execution after the instrumentation, 
which throws an exception when the main service is started. 
By applying \tool to the abnormal execution, we identify only 3 races, 
similarly to the report of RV~\cite{}.
In addition, the {\sf montecarlo} benchmark requires an external input 
that we downloaded from the internet and simplified. For the large 
applications, we utilized the most lightweight available configuration.

As the detection capability  depends on the observed run, for fair 
comparison we monitor the execution once by recording all necessary 
information required by both approaches, and then apply both techniques 
to the monitored run. Besides, our reported data for RV may be different 
from the original report for two reasons: (1) the different observed runs 
lead to different sets of races, (2) the original implementation of RV 
contains a bug in branch identification, which leads to  misses of 
many branches and  incorrect reduction of dependences during the analysis. 
We confirmed the bug with the author and fixed the bug in our 
experiments.\footnote{We have created an in-depth report on the bug, 
along with a witness test case. (See:  \url{https://sites.google.com/site/recipe3141/}.)}

Our experiments and measurements were all conducted on an x86 64 
Thinkpad W530 workstation with eight 2.30GHz Intel Core i7-3610QM, 
16GB of RAM and 6M cache. The workstation runs Ubuntu 12.04,
% of the Ubuntu Linux distribution, 
and has the Sun 64-Bit 1.6.0\_26 JVM installed.
%\footnote{We note that Sun JDK 1.7 does not support the transformation of dacapo applications with {\sf tamiflex}.}

@@

The Trace section includes the number of threads ($Th$), the number of 
shared reads ($Reads$) and shared writes ($Writes$), the subset of shared 
reads that read the base object references  ($Base$), where the  base object 
references are references used as the base/target in the following field 
reference or method invocation,  the number of branches ($Br$), the 
synchronization events ($Sync$) and the total number of events ($Total$), 
which includes local accesses in addition to the aforementioned events. 
Specifically, branch events are reported in the form $A/B$, where $A$ refers 
to the number of branches used by our analysis and $B$ refers to the number 
of branches used by RV. To ensure that the predicted run sees the same 
base object at each shared read/write, {\sf RV} inserts an artificial 
branch immediately in front of each shared access (and array accesses). 
We do not use such artificial branches.

We make a few interesting observations about the Trace section. 
The non-local events, i.e., all the events listed in Table~\ref{tab:main}, 
occupy around 30\% of the total trace in the first 7 benchmarks, 
but occupy less than 1\% of the trace in the rest of the benchmarks, 
which have relatively more complex logic. Reads of the shared base objects 
occupy a small portion ($\frac{1}{10}$-$\frac{1}{3}$) of the total 
shared reads. The rest of the shared reads read only primitive values 
or references not involved in field accesses (e.g., references involved 
in nullness checking). Besides, our analysis involves significantly less branch events as compared to the {\sf RV} approach. This difference plays an important role in the detection, which we will explain shortly.

@@

First of all, {\sf Recipe} finds $>$150 more races than {\sf RV}. There 
are several reasons for that: (1) {\sf Recipe} can reason about the 
accesses in the un-executed paths, while {\sf RV} and {\sf Recipe-s} can 
only reason about the accesses in the executed paths. (2) {\sf Recipe} 
or {\sf Recipe-s} relax the scheduling even if it 
%breaks the inter-thread read/write dependence structure 
changes variable values in the observed run. {\sf Recipe} allows the 
majority of the shared reads, i.e., the reads of primitive values, 
to freely read a different value from a different write 
as long as the value leads to the same branch decisions. {\sf RV}, however, 
requires them to read the same values as in the observed run. (3) An 
critical optimization proposed by {\sf RV} is to preserve dependences 
only for the reads before the preceding branches of the racy events, 
rather than all the reads. However, this optimization is underplayed 
by the fact that {\sf RV} introduces huge amount of artificial 
branches, i.e., one before each shared field access, which ensures 
the use of the same base objects.
We get rid of such artificial branches and instead, rely on the small 
amount of base read events to ensure the use of the same base 
objects (Section~\ref{}). Our strategy reduces the number of branches 
greatly and amplifies the effectiveness of the optimization. We also 
carried out case studies (Section~\ref{}) to better illustrate the scenarios.

Another interesting observation is that although {\sf Recipe} should 
produce all races found by {\sf RV} in theory, {\sf Recipe} may miss 
races found by {\sf RV} in practice (Column $Diff'$), i.e., {\sf Recipe} 
is not strictly more effective than {\sf RV} in practice. The reason  is 
the limitations of the underlying constraint solver: (1) the solver cannot 
compute constraints with very complex arithmatic operations, and (2) the 
solver does not support some program constants such as the scientific 
notation $3E-10$. In our empirical evaluation, we encounted only one 
benchmark, account, where these issues manifested.

The last section, Running time, compares the analysis time for both 
approaches. We find our approach is significantly slower than {\sf RV}, 
e.g., {\sf RV} often finishes within 200 seconds, while our approach may 
take more than 1 hour. This is because {\sf Recipe} has a larger search 
space and needs to reason about the computation among variables inside 
the local access events, while {\sf RV} needs to only reason about 
the order relations among the events.

@@

{\bf Relaxing Values\ } Figure~\ref{fig:relax1} demonstrates the relaxation of inter-thread dependence enabled by our approach. 
The code is from the benchmark {\sf bbuffer}, where the line numbers are
marked. Huang et al.~\cite{} detect the race between lines 291 and 400, 
but fail to detect the race between lines 294 and 400. The reason is 
that in the observed run, the execution follows the order 400, 291 and 294. 
For the event at line 294, its preceding branch at line 291 reads 
from line 400. Therefore, Huang et al.~\cite{}
require the predicted run to preserve the dependence between line 291 and 
line 400, so that line 291 reads exactly the same value and the branching 
expression evaluates to the same truth value. The dependence enforces 
the order constraint, $400 \rightarrow 291$, which further enforces 
the order $400\rightarrow 291 \rightarrow 294$.    Our approach does not 
require the existence of such dependence. Specifically, we allow line 
291 to happen before line 400 in the predicted run as long as the value 
read by it leads to the same branch decision, which is true in this case.  
As a result, there is no order constraint between line 294 and 
line 400, and the two form a race. We were able to replay this race  
easily using the eclipse IDE breakpoints.

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{figs/Visio-bbuffer.pdf}
\caption{Relaxation of Inter-thread dependence}\label{fig:relax1}
\end{figure}

{\bf Relaxing the Paths\ } Figure~\ref{fig:relax2} demonstrates how our 
approach relaxes the scheduling to account for  unexecuted paths in 
{\tt mergesort}. 
All threads invoke the method {\sf Sorting}, which recursively starts 
two child threads if there are two more available entries in the 
thread pool (lines 8-11), or starts one child thread if there is only 
one available entry (lines 4-5).  The availability is computed through 
the static method {\sf available} with the constant {\tt total}, which is equal to 5. The shared variable {\tt alive} denotes the used entries.

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{figs/Visio-msort.pdf}
\caption{Relaxation of Paths}\label{fig:relax2}
\end{figure}

\section{Threats to Validity}
A Java method can consist of up to 65535 bytecode instructions. Therefore,
 we count the number of bytecode instructions inside each method. If the 
resulting number exceeds 65525, we avoid instrumenting the method. 
The consequence is that we will potentially miss races inside a method for 
which we skipped instrumentation. In this case, we specify the variables 
read from the method to be equal to their concrete values in the constraints. 
The constraint solver can proceed safely without being affected by 
such methods.

\tool lacks support for certain operators. In our current 
implementation, we do not support bit operations such as $\&$ and $<<$, which accounts for most of our misses.
